---
title: "data.table: Swiss Army Knife for Data Wrangling"
author: "Paramjot Singh"
date: 2018-12-29T10:06:58-06:00
draft: FALSE
tags: ["R"]
categories: ["Tools"]
output:
  blogdown::html_page:
    toc: true
    toc_depth: 3
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  class.source = 'white',
  fig.align = 'center',
  eval = FALSE
)
```
While [data.table](https://github.com/Rdatatable/data.table/wiki) is mostly praised for its speed (which is undeniable), I was impressed with its versatility as well. The ability to manipulate the data in variety of ways using an extremely compact syntax is second to none. Sure it needs a bit of getting used to but the payoffs are phenomenal. I implemented one of my recent [project](https://github.com/pchhina/quizme) using data.table which helped me learn some of its really cool features. The package's [documentation](https://github.com/Rdatatable/data.table/wiki/Getting-started) does an impressive job too communicating the philosophy as well as usage. Here I document some of its features I have learned so far. I plan to update this regularly as I learn more.

I am going to use `ggplot2::mpg` data set for demonstration of its data wrangling features. So let's first create a data.table from this data:

```{r, eval = TRUE}
library(data.table)
mpg <- data.table(ggplot2::mpg)
mpg
```

### Row Selection
**Select first row**
```{r}
mpg[1]
```

**Select first three rows**
```{r}
mpg[1:3]
```

**Select all but first three rows**
```{r}
mpg[-(1:3)]
```

**Select rows five through seven but with order reversed**
```{r}
mpg[7:5]
```

**Select the last row**

`.N` is a special symbol to get the number of rows in the data table:
```{r}
mpg[.N]
```

**Select all rows except the last row**
```{r}
mpg[-.N]
```

**Random 10 observations**
```{r}
mpg[sample(.N, 10)]
```

**Random 10% observations**
```{r}
mpg[sample(.N, (.N * 0.1))]
```

**Top 10 rows**

Any data.frame functions can directly be used because data.table is also a data.frame
```{r}
head(mpg, 10)
```

**Bottom 10 rows**
```{r}
tail(mpg, 10)
```

**Select all `a4` model observations**
```{r}
mpg[model == "a4"]
```

**Select all `6` cylinder `a4` models**
```{r}
mpg[model == "a4" & cyl == 6]
```

**Select all non-`a4` model observations**
```{r}
mpg[model != "a4"]
```

**Select all models except `a4` and `passat`**
```{r}
mpg[!model %in% c("a4", "passat")]
```

**Select models with city fuel economy more than 25 mpg**
```{r}
mpg[cty > 25]
```

**Select every other row**
```{r}
mpg[rep(c(TRUE, FALSE), length = .N)]
```

**Select `compact` vehicles for `2008` model year**

Multiple variables can be used with `on` argument

```{r}
mpg[.(2008, "compact"), on = c("year", "class")]
```

**Amongst `subcompact` and `midsize` vehicles of `1999` model year, what is the best highway fuel economy**
```{r}
mpg[.(1999, c("subcompact", "midsize")), min(hwy), on = c("year", "class")]
```

**Select all observations with model name starting with *c* **
```{r}
mpg[grep("^c", model)]
```

**First row of `2008` model year `compact` vehicles**

`mult` argument can be used to get first/last observation in a sub-group

```{r}
mpg[.(2008, "compact"), mult = "first", on = c("year", "class")]
```

**Last row of `1999` model year `SUVs`**
```{r}
mpg[.(1999, "suv"), mult = "last", on = c("year", "class")]
```

**First row of all three subgroups: `4, 5 and 6` cylinder rear wheel drive cars**
```{r, eval = TRUE}
mpg[.(c(4, 5, 6), "r"), mult = "first", on = c("cyl", "drv")]
```

**First row of all three subgroups: `4, 5 and 6` cylinder rear wheel drive cars, but matching rows only**

`nomatch = 0L` can be used to show only matching rows

```{r, eval = TRUE}
mpg[.(c(4, 5, 6), "r"), mult = "first", nomatch = 0L, on = c("cyl", "drv")]
```

**First row of each class**

`.SD` can be used to select all columns

```{r}
mpg[, head(.SD, 1), by = class]
```

**Last row of each class**
```{r}
mpg[, tail(.SD, 1), by = class]
```

**Five worst highway fuel economy vehicles**
```{r}
head(mpg[order(hwy)], 5)
```

**Five best highway fuel economy vehicles**
```{r}
head(mpg[order(-hwy)], 5)
```

**Best fuel economy vehicle in each class**
```{r}
mpg[order(hwy), head(.SD, 1), by = class]
```

### Column (variable) Selection

**Select model, cyl, drv, hwy**
```{r}
mpg[, .(model, cyl, drv, hwy)]
```

**Select first three variables**
```{r}
mpg[, 1:3]
```

**Select all but first three variables**
```{r}
mpg[, -(1:3)]
```

**Select variables by first storing in a character vector**
```{r}
cols <- c("model", "cyl", "drv", "hwy")
mpg[, ..cols] # option 1
mpg[, cols, with = FALSE] # option 2
```

**Select all but variables stored in a character vector**
```{r}
cols <- c("model", "cyl", "drv", "hwy")
mpg[, -..cols] # option 1
mpg[, -cols, with = FALSE] # option 2
```

**Select all variables in the range displ:drv**
```{r}
mpg[, displ:drv]
```

**Select all variables except those in the range displ:drv**
```{r}
mpg[, -(displ:drv)]
```

**Select all variables starting with *m* **
```{r}
cols <- grep("^m", names(mpg))
mpg[, ..cols]
```

**Select all variables not starting with *m* **
```{r}
cols <- grep("^m", names(mpg))
mpg[, -..cols]
```

**Select all numeric columns **
```{r}
cols <- mpg[, sapply(.SD, is.numeric)]
mpg[, ..cols]
```

**Select all non-numeric columns **
```{r}
cols <- mpg[, sapply(.SD, is.numeric)]
mpg[, -..cols]
```

### Computing on J

**Number of observations**
```{r}
mpg[, .N]
```

**Number of observations in each class**
```{r}
mpg[, .N, by = class]
```

**Number of observations in each class, ordered alphabetically by class**
```{r}
mpg[, .N, keyby = class]
```

**Mean city fuel economy for each class**
```{r}
mpg[, mean(cty), by = class]
```

**Mean city fuel economy for each class and assign column name**
```{r}
mpg[, .(mean_city = mean(cty)), by = class]
```

**Mean city fuel economy for each class in increasing order**
```{r}
mpg[, .(mean_city = mean(cty)), keyby = class]
```

**Mean highway fuel economy for `honda` and `toyota` vehicles**
```{r}
mpg[manufacturer %in% c("honda", "toyota"), mean(hwy), by = manufacturer]
```

**Class of all variables, return as data table**
```{r}
mpg[, lapply(.SD, class)]
```

**Class of all variables, return as vector**
```{r}
mpg[, sapply(.SD, class)]
```

**Fraction of NA values for each variable**
```{r}
mpg[, lapply(.SD, function(x) mean(is.na(x)))]
```

**Mean of all numeric variables**

`.SDcols` can be used to apply a function to only subset of columns

```{r}
cols <- mpg[, sapply(.SD, is.numeric)]
mpg[, lapply(.SD, mean), .SDcols = cols]
```

**Change all non-numeric columns to upper case**
```{r}
cols <- mpg[, sapply(.SD, is.numeric)]
mpg[, lapply(.SD, toupper), .SDcols = -cols]
```

**Number of vehicles with average fuel economy more than 25**
```{r}
mpg[, sum(((cty + hwy) / 2) > 25)]
```

### More Expressions in J

**Scatterplot of city fuel economy versus displacement**
```{r, eval = TRUE}
mpg[, lattice::xyplot(cty ~ displ)]
```

**Density plot of highway fuel economy**
```{r, eval = TRUE}
mpg[, lattice::densityplot(~ hwy)]
```

### Update by Reference

`:=` is used to update data.table by reference

**Add a column for average fuel economy**
```{r, eval = TRUE}
mpg[, avg_mpg := (cty + hwy) / 2]
```

**Add a column with mean of city fuel economy**
```{r, eval = TRUE}
mpg[, avecty := mean(cty)]
```

**Add a factor variable (values = good, bad) based on city fuel economy being more or less than average city fuel economy**
```{r, eval = TRUE}
mpg[, mpgrating := ifelse(cty < mean(cty), "good", "bad")]
```

**Change 'f' to 'Front' for drv values**
```{r, eval = TRUE}
mpg["f", drv := "Front", on = "drv"]
```

### Remove Columns

**Remove `fl` column**
```{r, eval = TRUE}
mpg[, fl := NULL]
```

**Remove first three columns**
```{r, eval = TRUE}
mpg[, c(1:3) := NULL]
```

**Remove last column**
```{r, eval = TRUE}
mpg[, length(mpg) := NULL]
```

**Remove all columns starting with m**
```{r, eval = TRUE}
mpg[, grep("^m", names(mpg)) := NULL]
```

**Remove manufacturer and year columns**
```{r, eval = TRUE}
mpg[, `:=`(manufacturer = NULL,
           year = NULL)]
```

