---
title: "Persistent Plotting in R"
author: "Paramjot Singh"
date: "`r Sys.Date()`"
draft: FALSE
tags: ["R"]
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  class.source = 'white',
  fig.align = 'center',
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
library(ggplot2)
library(gridExtra)
library(themeSimple)
```
In this post, I describe the importance of persistent plots while carrying out a analysis. A case is made for simpler and persistent plots and the process to create the plots is demonstrated with an example. Finally, a package is created to stitch things together in a coherent fashion.

With the amount of available options in R today to create plots, it is easy to get distracted and miss the point of using plots in first place - to bring to attention something unique about the data which otherwise could be difficult, if not impossible, just looking at the raw numbers.

> *The simple graph has brought more information to the data analyst's mind than any other device.* - John Tukey

The approach I use here is to automate the theming process so that we do not have to think about it unless needed by:

* [Creating](../creating_theme) a new theme or modify the existing one to suit your needs.
* Using functions to modify plot related properties so these can be reused later.
* Using [package](../package_building) to make it easy to distribute or modify the stuff in future.
* Finally, automatically loading this package at the start of R session.

### Creating a theme

Creating a new theme is first step towards creating persistent plots. I have described the process in an earlier [post](../creating_theme) so I will not go into the details here.

### Using functions to modify plot related properties

This part is mostly driven by personal workflow since here we choose certain types of plots that we use more often and fine tune their look and feel. I am showing the default plots on the left and with my modifications on the right. I used the following plots:

1. Scatter plot: To show relationship between two continuous variables.

```{r scatter, out.width = "80%"}

s1 <- ggplot(mpg, aes(displ, hwy)) +
    geom_point() +
    theme_gray()

s2 <- ggplot(mpg, aes(displ, hwy)) +
    scatter() +
    theme_simple()

grid.arrange(s1, s2, nrow = 1)
```
2. Scatter plot with linear trendline: To show how the linear model fit looks like for a given set of continuous variables.

```{r fit_linear, out.width = "80%"}

fl1 <- ggplot(mpg, aes(displ, hwy)) +
    geom_point() +
    geom_smooth(method = "lm") +
    theme_gray()

fl2 <- ggplot(mpg, aes(displ, hwy)) +
    scatter_fit() +
    linear() +
    theme_simple()

grid.arrange(fl1, fl2, nrow = 1)
```
3. Scatter plot with smooth trendline: To show how the non-linear model fit looks like for a given set of continuous variables.

```{r fit_nonlinear, out.width = "80%"}

fn1 <- ggplot(mpg, aes(displ, hwy)) +
    geom_point() +
    geom_smooth() +
    theme_gray()

fn2 <- ggplot(mpg, aes(displ, hwy)) +
    scatter_fit() +
    nonlinear() +
    theme_simple()

grid.arrange(fn1, fn2, nrow = 1)
```

4. Histogram: To show frequency distribution of a single continuous variable.

```{r histogram, out.width = "80%"}

h1 <- ggplot(mpg, aes(hwy)) +
    geom_histogram() +
    theme_gray()

h2 <- ggplot(mpg, aes(hwy)) +
    histogram()  +
    theme_simple()

grid.arrange(h1, h2, nrow = 1)
```

5. Barplot: To show frequency distribution of a single discrete variable.

```{r bar, out.width = "80%"}

b1 <- ggplot(mpg, aes(as.factor(cyl))) +
    geom_bar() +
    theme_gray()

b2 <- ggplot(mpg, aes(as.factor(cyl))) +
    bar() +
    theme_simple()

grid.arrange(b1, b2, nrow = 1)
```

6. Boxplot: To show relationship between a continuous and discrete variable.

```{r box, out.width = "80%"}

x1 <- ggplot(mpg, aes(as.factor(cyl), hwy)) +
    geom_boxplot() +
    theme_gray()

x2 <- ggplot(mpg, aes(as.factor(cyl), hwy)) +
    box_simple() +
    theme_simple()

grid.arrange(x1, x2, nrow = 1)
```

### Building a package

I have already described this process in a post [here](../package_building), so I won't go into details.

### Automatically loading the theme

This involves creating a hook in your `.Rprofile` file. [Here](https://github.com/pchhina/dotfiles/blob/master/.Rprofile) is how my file looks like after making the changes. Now each time you start a new R session, the plot package (`themeSimple`) is automatically loaded into the workspace and you can use the helper functions to create the modified plots.
