---
title: "Quizme Introduction and Usage"
author: "Paramjot Singh"
date: 2018-10-08T19:04:48-05:00
draft: FALSE
tags: ["R"]
categories: ["Package"]
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  class.source = 'white',
  fig.align = 'center',
  warning = FALSE,
  message = FALSE,
  eval = FALSE
)
```

While writing code for our project, we often learn new functions and concepts. Unfortunately, we also forget the new knowledge quickly. Some of you might be lucky and remember the first time you read something. For the rest of us, we need deliberate practice. Wouldn't it be nice if we could practice any new material so we can recall it with ease? I have created a package, *quizme*, that can help you with that practice. You can build your personal quiz repository by adding questions and answers right from R console. When you want to practice, you can do that also right from the R console!

### How do I get the package?
Package is currently available on Github. You can download it using:
```{r, eval = FALSE}
library(devtools)
install_github("pchhina/quizme")
```
Then we will just load the library as usual:
```{r, eval = TRUE}
library(quizme)
```
**NOTE**: On Linux, I would recommend to also install [Autoquiz](https://github.com/pchhina/autoquiz) as it simplify the quiz workflow.

### How do I add questions?
Questions can be added with `addq()`. At the start of the session however, you have to use the function `quizme()`. This loads the previously created QA list or create blank data structures so new questions can be added.
```{r, eval = TRUE}
quizme()
```

With `addq()`, you can add questions in a single line. Answer must be added in a new line. Answer can contain multiple lines. Carriage return in a blank line will add your question-answer pair to an R object.
```{r}
addq()
```

### How do I take quiz?
You would use `ask()` and a randomly chosen question will be presented at the R console:
```{r, eval = TRUE}
ask()
```

### How do I look for the answer?
The answer to the question that was presented last can be seen using `tell()` function:
```{r, eval = TRUE}
tell()
```
### What if don't know the answer?
If you want to tell quizme me that you don't know the answer to the current question, you can use `miss()` function:
```{r}
miss()
```

### How many questions are remaining?
If you want to know the status of your quiz session, you can use `show_status()` function. In addition you could also use `week_ahead()` to see number of questions for each day for the next 7 days:
```{r}
show_status()
week_ahead()
```


### What do I do when I am done practicing?
Simply call `bye()`. This writes the updated QA data to a file in a directory that was created when `(quizme())` was called. After that, it also removes the data structures related to testing. And you are done. Happy learning!


### Learning More

- [Quizme Development](../quizme-development) for discussion of code and related functions
- [Quizme Design](../quizme-design) on discussion on design philosophy
- [Autoquiz Workflow](../quizme-workflow-with-autoquiz) for an example workflow using autoquiz
- [Quizme Package](https://github.com/pchhina/quizme) to install and try it!
- [Autoquiz](https://github.com/pchhina/autoquiz) to install and try autoquiz
